Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00

### 2019-01-23 manuscript.Rmd
rewrite conclusions

EiaD 2019 contains a large set of carefully curated, robustly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from GTEx. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues.  Furthermore, scRNA-seq datasets have been integrated, allowing for examination of cell type specific gene expression. As more human and non-human primate samples become publicly available, we will be updating the section eyeIntegration.

We have also provided a small example of the very close transcriptomic distance between fetal tissue - organoid - iPSC based retina. As human fetal tissue is extremely difficult to obtain, our dataset agnostically demonstrates the viability of using organoid and iPSC approaches to study human retina development.

Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates are substantially easier to add, and because all the processing is in modular pieces, and it is easier to add new analysis steps. In the future we plan on regularly adding new samples to EiaD, and also plan on more substantial steps like building *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic data.

>>>>

EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues. Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression. Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and once the raw data becomes publicly available, we will be updating the section eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].

As human fetal tissue is extremely difficult to obtain and not amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed. Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina developmental.

Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with minimal effort, and because all the processing is in modular pieces it is easier to add new analysis steps. In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic datasets.
